---
title: "lab_01"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your class reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the UMD courses data.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
umd_courses <- read_rds("lab_01.rds")
```

## Answer questions

**Q1** Which department has the highest mean (average) number of seats for its classes? You need to show the department with the highest average first. Which one is the most surprising/interesting to you? **First-Year Innovation & Research Experience with the third most mean of seats was the most interesting to me. This was because I was not aware of such to begin with and its surprising that this takes up so many seats.**

```{r}
highest_mean <- umd_courses |>
  group_by(department) |>
  summarise(
    mean_seats = mean(seats)
  ) |>
  arrange(desc(mean_seats))

highest_mean
```

**Q2** Find out the greatest number of times any individual class has been offered in our dataset. Your code must show the greatest number of classes first.
**Doctoral Dissertation Research is the class offered the most**

```{r}
class_count <- umd_courses |>
  group_by(title) |>
  summarise(
    class_count = n()
  ) |>
  arrange(desc(class_count))

class_count
```

**Q3** Count the number of courses for each instructor and department. Which instructor has taught the most?
**Keith Yearwood has taught the most classes with 79 classes. Anything before him did not list a name**

```{r}
instructor_top <- umd_courses |>
  group_by(department, instructors) |>
  summarise(
    class_count = n()
  ) |>
  arrange(desc(class_count))

instructor_top
```

**Q4** Given the results for Q3, what would have made it easier to find the answer? You don't have to know *how* to accomplish this; I want you to describe in words what would make the answer easier to see. How confident are you that the answer you got is the right one, and why? **Knowing how to leave out the NA's in the instructor category would help to get to the answer faster. Likely this would be with a filter selecting the NA's in the umd_courses. I'm about 85% confident in my response since I know that we were not taught how to remove NAs from the data set. **

**Q5** Take one of the questions you listed in the final question of the pre_lab and try to answer it here in code. If you don't know how to fully answer it, write what code you can and describe in words what else you would need to do. If you got an answer, is it interesting? **A5**

```{r}
# Which department has had the most growth in seat # and which department has had a decrease in the # of seats

seat_growth <- umd_courses |>
  group_by(term, department) |>
  summarise(
    classes_count = n()
  ) |>
  arrange(desc(classes_count))

seat_growth

# for the rest of the code I would need to implement percent change or just the general change between the different terms

# Before I do that, I would need to make a filter for terms / departments overtime to make it a time series 

```
