---
title: "lab_04"
author: "Derek Willis"
date: "2023-10-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)

# load your packages
library(tidyverse)
library(lubridate)
library(janitor)

#install.packages("openxlsx")

library(openxlsx)

#install.packages("refinr")

library(refinr)

```

For this lab, we want to investigate spending by Maryland state and local candidates during the 2022 election. For example, we could ask how much money have candidates and committees spent on advertising on Facebook? Which vendors received the most money for media expenses? We have the data, but it's messy - names of payees are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of Maryland campaign expenditures located in the data folder called "maryland_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to clean up the column names and make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table

md_expenditures <- read_csv("data/maryland_expenses.csv")

md_expenditures <- md_expenditures |>
  clean_names() |>
  mutate(expenditure_date = mdy(expenditure_date))

```

## Answer questions

**Q1.** You've been assigned to write about Maryland political committees' use of Facebook advertising. There's a category column in the data that has values for different kinds of spending. One of them is "Media". You'll need to complete the following steps:

1.  Create a dataframe of Media expenses where the name of the payee is not NA.
2.  Then, using `write_csv`, make a CSV file of your media expenses in your data folder. `write_csv()` is similar to `read_csv()` but you provide it with the dataframe you're writing and the file path.
3.  Using OpenRefine, create a project using that CSV file, create a copy of the `payee_name` column called `payee_clean` and then standardize it, focusing on the payees that appear most often. Pay particular attention to variations of Facebook, bearing in mind that it changed its name in 2021: <https://about.fb.com/news/2021/10/facebook-company-is-now-meta/>), but also clean up variations of the most common payees.
4.  When you are done standardizing `payee_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.
5.  Using that `payee_clean` column in your new dataframe, calculate the total amount of Facebook spending by each committee and present the results showing the committee that spent the most first.

**A1.**

```{r}

no_nas_md_expenses <- md_expenditures |> 
  drop_na(payee_name)

#write.xlsx(no_nas_md_expenses, "data/no_nas_md_espenses.xlsx")


# Reading back in
 refined_md_expenses <- read_csv("data/cleaned-maryland-expenses-refined.csv")

top_spendors <- refined_md_expenses |>
  filter(expense_category == "Media") |>
  group_by(committee_name) |>
  mutate(payee_clean = str_to_lower(payee_clean)) |>
  filter(
    str_detect(payee_clean, "facebook") |
    str_detect(payee_clean, "meta | meta ")
    ) |>
  summarise(
    total_spent = sum(amount)
  ) |>
  arrange(desc(total_spent))

top_spendors

```

**Q2.** Let's do some research on the top committees from question 1. Look up the top 10 spenders: if they are candidate committees, what offices were they running for? If they were not candidates, what issues do they seem to care about? Write some sentences answering those questions, then go the Facebook ad library ([https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data\\[direction\\]=desc&sort_data\\[mode\\]=relevancy_monthly_grouped&media_type=all](https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data%5Bdirection%5D=desc&sort_data%5Bmode%5D=relevancy_monthly_grouped&media_type=all){.uri}) and look for ads from the top spender. What do you notice about the the information provided about those ads?

**Top spenders: 1) Jain Ashwani Friends of: Candidate committee and was running for Maryland House District 6 2) Governor candidate 3) Non-candidate, does not seem to be advocating for anything in particular 4) Non-candidate but does not talk about anything politically other than their name: Working Families Party 5) Non-candidate, no discussion of politics 6) Governor candidate 7) Governor candidate 8) Non-candidate, unknown politically 9) Non-candidate, pro-fire union 10) Non-candidate, politically unknown . . . its gets you a spcific id, the date that it ran for, the platforms it was shown on, categories and estimated audience size, amount spent and impressions . . . along with add itself. **

**Q3.** Given the list of top spenders in A1 and the primary results for governor (<https://elections.maryland.gov/elections/2022/primary_results/gen_results_2022_1.html>), who is missing from that Q1 list? Did those campaigns run Facebook ads? Who are the biggest recipients of media expenses by the Democratic primary winner for governor (HINT: that person has two committees, the second one with his running mate)?

**1) The main candidates are missing for example Dan Cox and Gordana Schifanelli / Wes Moore and Aruna Miler. 2) It does not see like the big two ran facebook ads or at least spent money on it. 3) SKDK was given the largest total recipients of media expenses: 83 media expenses recived.**

```{r}

top_spendors |>
  filter(
    str_detect(str_to_upper(committee_name), "WES|MOORE") |
    str_detect(str_to_upper(committee_name), "ARUNA|MILER") |
    str_detect(str_to_upper(committee_name), "DAN|COX") |
    str_detect(str_to_upper(committee_name), "GORDANA|SCHIFANELLI")
    )

ads_with_facebook <- refined_md_expenses |>
  filter(
    str_detect(str_to_lower(committee_name), "moore|wes ") &
    str_detect(str_to_lower(committee_name), "aruna|miller") |
    str_detect(str_to_lower(committee_name), "moore|wes ") |
    str_detect(str_to_lower(committee_name), "cox|dan ") &
    str_detect(str_to_lower(committee_name), "gordana|schifanelli") |
    str_detect(str_to_lower(committee_name), "cox|dan ") &
    str_detect(str_to_lower(committee_name), "governor")
  ) |>
  filter(
    str_detect(str_to_lower(payee_clean), "facebook") |
    str_detect(str_to_lower(payee_clean), "meta | meta ")
  )

ads_with_facebook

moore_media_expenses <- refined_md_expenses |>
  filter(
    str_detect(str_to_lower(committee_name), "moore|wes ") &
    str_detect(str_to_lower(committee_name), "aruna|miller") |
    str_detect(str_to_lower(committee_name), "moore|wes "),
    expense_category == "Media"
  ) |>
  group_by(payee_clean) |>
  summarise(
    number_of_media_expenditures_received = n()
  ) |>
  arrange(desc(number_of_media_expenditures_received))
  

moore_media_expenses

```

**Q4.** The media expense data you cleaned also has a standardized purpose column. Write code to generate a list of the committees with the total amount of spending on online advertising. How does this list compare to the one from Q1 about top Facebook spenders? Of the top 5 committees who spent the most on Facebook ads from Q1, are there any that appear to have spent *all* of their online advertising dollars on Facebook?

**they differ greatly, with all of the committees generated different from Q1 Jain Ashwani Friends of and Frenchot Anderson-Walker for Maryland Slate spent all their online advertising money on facebook.**

```{r}

md_expenditures |>
  filter(str_to_lower(expense_purpose) == "online advertising" ) |>
  group_by(committee_name) |>
  summarise(
    total_amount_spent = sum(amount)
  ) |>
  arrange(committee_name)


```

**Q5** Your editor asks you to summarize your findings about spending on Facebook by Maryland political candidates. How much do we know and what don't we know? What's the best story idea you've found here?

**We know a great deal about the smaller campaigns. Those who ran smaller more grassroots campaigns advertised largely on Facebook while larger campaigns did not seem to post such ads on Facebook. We know that some of the campaigns spent their entire online advertisment expenditures on Facebook and not any other platform. We do not have a comprehensive list of the other places that the smaller groups advertised with. The main story idea is the difference in online presence for smaller vs larger campaigns.**
